Results

### 2022-07-11 article/methods.Rmd
first poster draft

---
title: "Methods"
output: 
  # html_notebook
  bookdown::html_document2:
    number_sections: FALSE
bibliography: 
- ../../library.bib
- ../../grateful-refs.bib
link-citations: TRUE
---

@@

```{r map, fig.width=4, fig.height=4, fig.cap="Map of conservation areas and study site on Osa Peninsula, Costa Rica. Map image data © 2022 Osa Conservation. Photo credit: Nicholas Medina."}
source(here::here("figs/map.R"));
sitePlot
# knitr::include_graphics(here::here("figs/figmap.png"))
# magick::image_read(path = "../figs/map.png") %>% plot()
```
<!-- ![**Fig 1**](../figs/map.png) -->

@@

The diameter of all stems >10 cm wide were recorded in each census plot.
Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem. 
Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.
Taxa were identified with localized knowledge and field guides matching common names to Latin binomials, 
and trait information was gathered from the literature.

@@

Aboveground biomass (AGB) was estimated per stem using the allometric equation by *@chave15*.
The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *@chave05 [@alvarez12; @ngomanda14]*.
The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *[@domke12; @khan20]*.
Wood density (or specific gravity) values when available also improve biomass estimates even more than height *[@vanbreugel11]*.
The equation used was: 
$${AGB} = 0.0673 (p D^2 H)^{0.976}$$ 
with diameter *D* in cm, height *H* in m, and wood density or specific gravity *p* in *g cm^-3^*.
DBHs and heights were measured in the field and wood densities were extracted from available literature. 
Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be *0.58* following World Agroforestry Database guidelines [(worldagroforestry.org)](worldagroforestry.org).

<!-- *[References used for this step are given in *Table S1*.] * -->

Original values found and recorded were later updated to modern values using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by *@rejou-mechain17*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.

<!-- However, these updates did not appear to change the main results of this study. -->

@@

For all response variables, data were analyzed using plot medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base `r version$version.string` and *rstatix* `r packageVersion("rstatix")` package functions *[@base]*.
 <!-- including biomass converted to a per hectare basis,  -->

<!-- Correlated variables were grouped into multi-way ANOVAs---namely biomass, wood density, and diameter in one group; and height, stem density, and wood density in a different group.  -->
Non-linear regressions were run using the function *poly()* in the base R package *stats* `r packageVersion("stats")`, 
which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models. 
All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.
Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and are available at [nmedina17.github.com/osa](nmedina17.github.com/osa).

>>>>

---
title: "Methods"
output: 
  # html_notebook
  bookdown::html_document2:
    number_sections: FALSE
bibliography: 
- ../../library.bib
- ../../grateful-refs.bib
- ../../software.bib
link-citations: TRUE
---

@@

```{r map, fig.width=4, fig.height=4, fig.cap="Map of (a) conservation areas and (b) study site of (c) Bombacopsis quinata timber plantation on Osa Peninsula, Costa Rica. Map image data © 2022 Osa Conservation. Photo credit: Nicholas Medina."}
source(here::here("figs/map.R"));
sitePlot
# knitr::include_graphics(here::here("figs/figmap.png"))
# magick::image_read(path = "../figs/map.png") %>% plot()
```
<!-- ![**Fig 1**](../figs/map.png) -->

@@

The diameter of all stems >10 cm wide were recorded in each census plot, totaling over 1,000 trees.
Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem. 
Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.
Taxa were identified with localized knowledge and field guides matching common names to Latin binomials, 
and trait information was gathered from the literature.

@@

Aboveground biomass (AGB) was estimated per stem using the allometric equation by *@chave15*.
The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *@chave05 [@alvarez12; @ngomanda14]*.
The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *[@domke12; @khan20]*.
Wood density (or specific gravity) values when available also improve biomass estimates even more than height *[@vanbreugel11]*.
The equation used was: 
$${AGB} = 0.0673 (p D^2 H)^{0.976}$$ 
with diameter *D* in cm, height *H* in m, and wood density or specific gravity *p* in *g cm^-3^*.
DBHs and heights were measured in the field and wood densities were extracted from available literature. 
Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be *0.58* following World Agroforestry Database guidelines [(worldagroforestry.org)](worldagroforestry.org).
<!-- *[References used for this step are given in *Table S1*.] * -->
Original values found and recorded were later updated to modern values using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by *@rejou-mechain17*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations. 
<!-- However, these updates did not appear to change the main results of this study. -->

Traits including successional stage and main dispersal mode were matched to species with *@werden20* data, and wood tissue carbon matched to genera with the GLOWCAD database *[@doraisami22]* given high alignment at species levels *[@chave06]*.

@@

For all response variables, data were analyzed using plot or distance strata medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base `r version$version.string` and *rstatix* `r packageVersion("rstatix")` package functions *[@base]*.
 <!-- including biomass converted to a per hectare basis,  -->

<!-- Correlated variables were grouped into multi-way ANOVAs---namely biomass, wood density, and diameter in one group; and height, stem density, and wood density in a different group.  -->
Non-linear regressions were run using the function *poly()* in the base R package *stats* `r packageVersion("stats")`, 
which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models. 
All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.
Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22]* are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa).

### 2022-07-11 article/results.Rmd
first poster draft

Overall diversity changed significantly with distance to primary forest edge 
(Fig \@ref(fig:taxa)).
Alpha diversity significantly 
(`r glue("p=",round(as.numeric(entropy12$pval[[2]][2]),3))`) 
decreased non-linearly by 
~`r abs(entropy12$statPrint[[1]]$estimate[[2]])` ± 
`r abs(entropy12$statPrint[[1]]$std.error[[2]])` per m 
as distance to primary forest edge increased, 
which explained 
`r "R2x100"`% of variance 
among distance strata medians 
(Fig \@ref(fig:taxa)a) -- 
while in contrast taxonomic richness increased slightly 
(`r glue("p=",round(as.numeric(richness$pval),3))`) 
and linearly 
by 
`r richness$statPrint[[1]]$estimate[[2]]` ± 
`r richness$statPrint[[1]]$std.error[[2]]` per m 
and had `r "R2x100"`% variance explained 
among distance strata 
(Fig \@ref(fig:taxa)b).
Community composition and beta diversity also changed significantly 
(`r glue("p=",ordStat[[5]][1])`) with 
`r ordStat$R2[1] *100`% 
variance among distance strata explained by distance to primary forest edge, 
and the first two principal components explaining 
`r ordTbl$PC1_prop[1] *100`% and 
`r ordTbl$PC2_prop[1] *100`% totaling 
`r (ordTbl$PC1_prop[1] + ordTbl$PC2_prop[1]) *100`% 
of variance among calculated plot distances 
(Fig \@ref(fig:taxa)d).
Key abundant taxa *`r measure2e`* and *`r measure2f`* showed different responses -- *`r measure2e`* nearly tended to decrease with distance to primary forest edge 
(Fig \@ref(fig:taxa)e),
while *`r measure2f`* decreased marginally significantly 
(`r glue("p=",round(as.numeric(ficus$pval),3))`) 
and linearly by 
`r ficus$statPrint[[1]]$estimate[[2]]` ± 
`r ficus$statPrint[[1]]$std.error[[2]]` per m 
with `r "R2x100"`% variance among distance strata explained 
(Fig \@ref(fig:taxa)f).

@@

## Functions

@@

Overall functional regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest 
(Fig \@ref(fig:traits)).
Taxa associated with both early and late successional stages decreased marginally significantly 
(`r glue("p=",round(as.numeric(succession$pval[[7]][2]),3))`) 
and curvi-linearly 
by 
~ `r succession$statPrint[[7]]$estimate[[2]]` ± 
`r succession$statPrint[[7]]$std.error[[2]]` kg per m 
with distance to edge explaining 
`r "R2x100"`% variance 
among distance strata medians 
(Fig \@ref(fig:traits)a).
Dispersal modes did not show consistent trends in biomass with increasing distance to edge 
(Fig \@ref(fig:traits)b).

>>>>

Overall diversity changed significantly with distance to primary forest edge 
(Fig \@ref(fig:taxa)).
Diversity significantly 
(`r glue("p=",round(as.numeric(entropy12$pval[[2]][2]),3))`) 
decreased non-linearly by 
~`r abs(entropy12$statPrint[[1]]$estimate[[2]])` ± 
`r abs(entropy12$statPrint[[1]]$std.error[[2]])` per m 
as distance to primary forest edge increased, 
which explained 
`r "R2x100"`% of variance 
among distance strata medians 
(Fig \@ref(fig:taxa)a) -- 
while in contrast taxonomic richness increased slightly 
(`r glue("p=",round(as.numeric(richness$pval),3))`) 
and linearly 
by 
`r richness$statPrint[[1]]$estimate[[2]]` ± 
`r richness$statPrint[[1]]$std.error[[2]]` per m 
and had `r "R2x100"`% variance explained 
among distance strata 
(Fig \@ref(fig:taxa)b).
Community composition and beta diversity also changed significantly 
(`r glue("p=",ordStat[[5]][1])`) with 
`r ordStat$R2[1] *100`% 
variance among distance strata explained by distance to primary forest edge, 
and the first two principal components explaining 
`r ordTbl$PC1_prop[1] *100`% and 
`r ordTbl$PC2_prop[1] *100`% totaling 
`r (ordTbl$PC1_prop[1] + ordTbl$PC2_prop[1]) *100`% 
of variance among calculated plot distances 
(Fig \@ref(fig:taxa)d).
Key abundant taxa *`r measure2e`* and *`r measure2f`* showed different responses -- *`r measure2e`* nearly tended to decrease with distance to primary forest edge 
(Fig \@ref(fig:taxa)e),
while *`r measure2f`* decreased marginally significantly 
(`r glue("p=",round(as.numeric(ficus$pval),3))`) 
and linearly by 
`r ficus$statPrint[[1]]$estimate[[2]]` ± 
`r ficus$statPrint[[1]]$std.error[[2]]` per m 
with `r "R2x100"`% variance among distance strata explained 
(Fig \@ref(fig:taxa)f).

@@

## Traits

@@

Overall trait regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest 
(Fig \@ref(fig:traits)).
Taxa associated with both early and late successional stages decreased significantly 
(`r glue("p=",round(as.numeric(succession$pval[[7]][2]),3))`) 
and curvi-linearly 
by 
~ `r succession$statPrint[[7]]$estimate[[2]]` ± 
`r succession$statPrint[[7]]$std.error[[2]]` kg per m 
with distance to edge explaining 
`r "R2x100"`% variance 
among distance strata medians 
(Fig \@ref(fig:traits)a).
Dispersal modes did not show consistent trends in biomass with increasing distance to edge 
(Fig \@ref(fig:traits)b).
